Counterfactual Training

### 2024-12-18 paper/presenations/slides.qmd
presentation for friday

Combine ideas from Energy-Based Models and Adversarial Training:

@@

- ${x^\prime}_t$ are counterfactuals of $x_{s}\subseteq x$ with target class $t$.
- $\ell_{\text{gen}}$ is the difference in energies between observed samples and counterfactuals.
- Counterfactuals are recycled as adversarial examples.

@@

**Example**: Let $f_\theta(x)=\theta^{T}x$ be a linear classifier. Suppose that featue $x_1$ is immutable (e.g. 'gender' or 'faith').

$$
\begin{aligned}
\nabla_\theta \ell_{\text{gen}}(x^\prime,x;\theta) &= \nabla_\theta ( \theta^{T}x[t] - \theta^{T}x^\prime[t]) \\
\frac{\partial \ell_{\text{gen}}}{\partial\theta_1}(x^\prime,x;\theta)&=x_1 - x_1^\prime
\end{aligned}
$$

- If $$

>>>>

Counterfactual Training (CT) combines ideas from Energy-Based Models and Adversarial Training:

@@

- ${x^\prime}_t$ are counterfactuals of $x_{s}\subseteq x$ with target class $t$.
- $\ell_{\text{gen}}$ is the difference in energies between observed samples in target class $x_t$ and counterfactuals.
- Counterfactuals are recycled as adversarial examples.

@@

Let $f_\theta(x)=\theta^{T}x$ be a linear classifier:

$$
\begin{aligned}
\nabla_\theta \ell_{\text{gen}}({x^\prime}_t,x_t;\theta) &= \nabla_\theta ( \theta^{T}x_t - \theta^{T}{x^\prime}_t) \\
\frac{\partial \ell_{\text{gen}}}{\partial\theta[1]}(x^\prime,x;\theta)&=x_{t}[1] - x_{t}^\prime[1]
\end{aligned}
$$

Suppose that featue $x[1]$ is immutable (e.g. 'age'), so $x_{t}^\prime[1]=x_s[1]$ where $s \neq t$. If $x_{t}[1] > x_s[1]$:

- $\ell_{\text{gen}}$ induces lower values of $\theta[1]$, acting as a hedge against $\ell_{\text{clf}}$, which favours higher $\theta[1]$.

# Findings

## Moons (Plausibility)

- All counterfactuals at test time generated using *ECCo*.
- Penalty on energy differential increases from *l.* to *r.*

![Plausibility of faithful counterfactuals $x_{t}^\prime$ measured in terms of their distance from $x_{t}$.](www/moons_plaus.png){#fig-moons-plausib}

## Moons (Example)

::::{.columns}::::
:::{.column width='50%'}
![Counterfactual explanations for model trained with CT (*ECCo*).](www/moons_ce_ecco.png){#fig-ce-ecco}
:::
:::{.column width='50%'}
![Counterfactual explanations for conventionally trained model.](www/moons_ce_vanilla.png){#fig-ce-vanilla}
:::
::::

## Moons (Validation Accuracy)

![Validation accuracy for different models.](www/moons_acc_val.png){#fig-moons-acc-val}

# Planning Ahead
